---
title: "COVID-19 Final Project"
author: "Rophence Ojiambo"
date: last-modified
lightbox: true
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    theme:
      light: flatly
      dark: darkly
    highlight-style: github
link-citations: true
colorlinks: true
linkcolor: blue
urlcolor: blue
bibliography: references.bib
csl: apa-numeric-superscript-brackets.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      dpi = 300)
```

::: panel-tabset
# Introduction

## Overview

The novel Coronavirus disease (COVID-19), caused by the serious acute respiratory syndrome (SARS-CoV-2) [@synowiec2021severe] virus is a highly infectious respiratory disease, first identified in Wuhan, China, in December 2019 [@zhu2020novel]. Since its emergence, the pandemic has impacted millions worldwide, with over 770 million confirmed cases and approximately 7 million deaths reported globally by the [**World Health Organization’s (WHO) Coronavirus (COVID-19) Dashboard**](https://data.who.int/dashboards/covid19/cases?n=c). The COVID-19 pandemic, which was first traced in the United States in January 2020 [@holshue2020first], highlighted and worsened existing health disparities, with racial and ethnic minority groups being disproportionately affected. Communities such as Black, Hispanic, Asian, and Native American populations faced higher rates of infection [@hooper2020covid], hospitalizations [@garg2020hospitalization; @romano2021trends], and mortality compared to White populations [@rossen2020excess; @gold2020race; @tai2021disproportionate]. Understanding the patterns and magnitude of these disparities is critical for informing targeted public health interventions for future pandemics.

## Problem Statement

This project aims to investigate the disproportionate impact of the COVID-19 pandemic on racial and ethnic communities in the United States. By analyzing disparities in cases, deaths, hospitalizations, and testing rates across states and racial/ethnic groups, the research seeks to illuminate the differential burden of COVID-19 on these communities. The insights derived from this study will be vital for public health planning and policy making to promote equity and preparedness for future public health emergencies.

## Data and Methodology

To explore these disparities, I will utilize the following datasets:

-   **Nation level: us.csv**
-   **State level: us-states.csv**

The above two datasets contain the US COVID-19 data (cumulative cases and deaths) till 03/23/2023, and were originally obtained from [The New York Times](https://www.nytimes.com/interactive/2021/us/covid-cases.html), [Github repository](https://github.com/nytimes/covid-19-data)

Additionally, I will use the following external data sets to improve the analysis on the required data:

-   [**COVID Racial Data Tracker**](https://covidtracking.com/race): Provides state-level information on cumulative COVID-19 cases and deaths (overall and by racial/ethnic group), as well as the number of hospitalizations and tests conducted by racial/ethnic group.
-   [**American Community Survey (ACS) Demographic and Housing 5-Year Estimates (2019)**](https://data.census.gov/table/ACSDP5Y2019.DP05): Contains U.S. population and state-level population estimate data for 2019, enabling the calculation of rates for cases, deaths, hospitalizations, and testing across racial and ethnic groups.

These datasets will be merged to facilitate a comprehensive analysis of state-level disparities.

## Proposed Analytic Approach

The analysis will involve the following steps:

1.  **Exploratory Data Analysis:**

-   Data cleaning, transformation of variables, missing data,

2.  **Descriptive Analysis and Data Visualization:**

-   Overall description of Total Covid Cases and Deaths
-   Calculate national and state-level rates for COVID-19 cases, deaths, hospitalizations, and tests.
-   Use visualizations such as bar charts and maps to compare these counts/rates across states and changes over time.

3.  **Correlation Analysis:**

-   COVID-19 cases, deaths, hospitalizations, and tests by racial and ethnic groups and changes over time as the pandemic progressed.
-   Fit a multiple linear regression model to examine the relationship between race, number of tests, number of hospitalizations and COVID-19 deaths.

## Contribution

This analysis will shed light on how the COVID-19 pandemic disproportionately affected racial and ethnic minorities in the United States. By examining and comparing the differences in cases, deaths, and hospitalizations across populations and states, this project will highlight areas where inequities are most severe. The findings can help leaders and public health officials focus their efforts, distribute resources fairly, and address the underlying issues that make some communities more vulnerable during health crises.

# Packages Required

Next, we load all the packages required for the analysis

```{r}
# Loading required libraries

library(tidyverse)   # A collection of R packages for data manipulation, visualization, and analysis
library(knitr)
library(kableExtra)
library(DT)
library(glue)        # For creating and combining strings with embedded variables or expressions
library(lubridate)   # For working with dates and times in a user-friendly way
library(stringr)     # For handling string operations like pattern matching, substitution, etc.
library(naniar)      # For managing and visualizing missing data in datasets
library(usdata)      # For working with datasets specific to the United States, including state and county data
library(ggplot2)     # For creating and customizing data visualizations
library(htmltools)

```

# Data Preparation - 10 points

## Original Source

### National-level and State-level data

The national-level (`us.csv`) and state-level (`us-states.csv`) COVID-19 datasets were sourced from [The New York Times](https://www.nytimes.com/interactive/2021/us/covid-cases.html) and are publicly accessible through their [GitHub repository](https://github.com/nytimes/covid-19-data). These datasets document cumulative COVID-19 cases and deaths across the United States up to March 23, 2023. Compiled by a dedicated team of journalists, the data was gathered by monitoring press conferences, analyzing public health reports, and seeking clarification from officials to ensure accuracy and completeness. This effort addressed challenges in a fragmented public health system, where overwhelmed state, county, and territorial agencies often struggled with consistent and timely reporting. The datasets include both confirmed and probable cases and deaths, adhering to definitions established by the Council of State and Territorial Epidemiologists and the Centers for Disease Control and Prevention (CDC). Corrections or reassignments between counties or states sometimes led to declines in reported cases or deaths on certain dates. The case counts represent all cases, including individuals who recovered or died. Updated daily, the datasets reflect real-time changes, with retroactive corrections applied when errors were identified or new information became available. To preserve transparency, the datasets retain their raw reported values with no imputed values.

#### Variables Overview

1.  **`us.csv` (National-Level Data)**: has 1,158 observations and 3 variables (`date`, `cases`, `deaths`)
2.  **`us-states.csv` (State-Level Data)**: has 61,942 observations and 5 variables (`date`, `state`, `fips` (Federal Information Processing Standards (FIPS) code for geographic identification), `cases`, `deaths`)

### American Community Survey (ACS) Demographic and Housing 5-Year Estimates (2019)

The data for this analysis was obtained from the [U.S. Census Bureau's American Community Survey (ACS) 5-Year Estimates](https://data.census.gov/table/ACSDP5Y2019.DP05?t=Populations%20and%20People&g=010XX00US_040XX00US01,02,04,05,06,08,09,10,11,12,13,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,44,45,46,47,48,49,50,51,53,54,55,56,60,66,69,72,78). The Census Bureau's [Population Estimates Program (PEP)](https://www.census.gov/programs-surveys/popest/about.html) produces estimates of the population for the United States, its states, counties, cities, and towns, as well as for Puerto Rico. PEP annually utilizes current data on births, deaths, and migration to calculate population change since the most recent decennial census and produce a time series of estimates of population, demographic components of change, and housing units. These estimates are used in federal funding allocations, as survey controls, as denominators for vital rates and per capita time series, and as indicators of recent demographic changes. The ACS 5-Year Estimates (2019) include data collected from January 1, 2015, to December 31, 2019. As a sample survey, the ACS provides estimates accompanied by margins of error (MOE).

#### Variables Overview

-   **Label**: Represents the name of the geographic region (e.g., United States, individual states).
-   **Total population**: The total population estimate for each geographic region.
-   **Two or more races**: Population estimate for individuals who identify with two or more racial groups.
-   **White**: Population estimate for individuals identifying as White.
-   **Black or African American**: Population estimate for individuals identifying as Black or African American.
-   **American Indian and Alaska Native**: Represents individuals identifying as Native American or Alaska Native.
-   **Asian**: Population estimate for individuals identifying as Asian.
-   **Native Hawaiian and Other Pacific Islander**: Represents individuals identifying as Native Hawaiian or Pacific Islander.
-   **Hispanic or Latino (of any race)**: Population estimate for individuals identifying as Hispanic or Latino, regardless of their racial category.
-   **Not Hispanic or Latino**: Population estimate for individuals identifying as non-Hispanic.

### COVID Racial Data Tracker data

The CRDT (COVID Racial Data Tracker) data, labelled `CRDT Data - CRDT.csv` was originally obtained from the [COVID Racial Data Tracker](https://covidtracking.com/race/about), a project developed in partnership with The [Atlantic’s COVID Tracking Project](https://covidtracking.com/) and the [Boston University Center for Antiracist Research](https://www.bu.edu/antiracism-center/), founded by Activist and scholar Dr. Ibram X. Kendi. This dataset provides information on the racial and ethnic disparities in COVID-19 cases, deaths, hospitalizations, and tests conducted across 51 states/territories, meaning granular insights at the county or local levels are not directly available.. The data is publicly available and was accessed from the COVID Racial Data Tracker website, which served as a key resource during the pandemic to highlight inequities in COVID-19 outcomes. The original dataset, which can be directly using this [link](https://docs.google.com/spreadsheets/d/e/2PACX-1vS8SzaERcKJOD_EzrtCDK1dX1zkoMochlA9iHoHg_RSw3V8bkpfk1mpw4pfL5RdtSOyx_oScsUtyXyk/pub?gid=43720681&single=true&output=csv), includes **5,320 rows** and **54 variables**, covering data collected from **April 12, 2020**, to **March 7, 2021**, across various states/territories in the United States. The Racial Data Tracker aimed to collect racial and demographic data to understand the outbreak and protect vulnerable communities. In the original dataset, missing values were represented as blank cells. Upon loading the data into R, these blanks were automatically interpreted as `NA`. These missing entries occur because not all states reported COVID-19 data disaggregated by race or ethnicity consistently during the pandemic.

#### Variables Overview

1.  **Date and State**:
    -   `Date`: The reporting date in the format `YYYY-MM-DD`.
    -   `State`: The U.S. state or territory where data was reported.
2.  **COVID-19 Outcomes**:
    -   **Cases**: Total and race/ethnicity-specific cumulative COVID-19 cases (e.g., `Cases_Total`, `Cases_White`, `Cases_Black`).
    -   **Deaths**: Total and race/ethnicity-specific cumulative deaths due to COVID-19 (e.g., `Deaths_Total`, `Deaths_White`, `Deaths_Black`).
    -   **Hospitalizations (`Hosp_`)**: Total and race/ethnicity-specific cumulative hospitalizations for COVID-19 (e.g., `Hosp_Total`, `Hosp_Black`).
    -   **Tests (`Tests_`)**: Total and race/ethnicity-specific COVID-19 tests conducted (e.g., `Tests_Total`, `Tests_Latinx`).

##### Racial and Ethnic Groups

The dataset categorizes individuals into the following racial and ethnic groups:

| **Variable** | **Description** |
|------------------|-----------------------------------------------------|
| **White** | Individuals identified as non-Hispanic White. |
| **Black** | Individuals identified as non-Hispanic Black or African American. |
| **Latinx** | Individuals identified as Hispanic or Latino. |
| **Asian** | Individuals identified as non-Hispanic Asian. |
| **AIAN** | Individuals identified as non-Hispanic American Indian or Alaska Native. |
| **NHPI** | Individuals identified as non-Hispanic Native Hawaiian or Pacific Islander. |
| **Multiracial** | Individuals identified as belonging to multiple racial groups. |
| **Other** | Individuals reported in categories not otherwise specified. |
| **Unknown** | Individuals whose racial or ethnic data were not reported or categorized. Missing values were left blank and interpreted as `NA` in R. |
| **Hispanic Ethnicity** | Individuals identified as Hispanic Ethnicity |
| **Non-Hispanic Ethnicity** | Individuals identified as Non-Hispanic Ethnicity. |
| **Ethnicity Unknown** | Individuals whose ethnicity was not recorded. |

: Racial and Ethnic Groups in the Dataset {#tbl-racial-ethnic-groups}

## Data Importing and Cleaning

### Processing and Cleaning the state-level and national-level, and population estimates Data sets

Four datasets were loaded for analysis: `us-states.csv` (state-level COVID-19 data), `us.csv` (national-level COVID-19 data), `CRDT Data - CRDT.csv` (COVID Racial Data Tracker), and `ACSDP1Y2019.DP05-2024-11-18T022938.csv` (American Community Survey population data). Each dataset was checked for duplicate rows using the `duplicated()` function, and no duplicates were found in any of the datasets.

```{r, include=FALSE}
################################################################################
#                  DATA IMPORTING & CHECKING FOR DUPLICATES                    #
################################################################################ 

# Load the datasets

states_data <- read_csv("Data/us-states.csv")
us_data <- read_csv("Data/us.csv")
covid_racial_data <- read_csv("Data/CRDT Data - CRDT.csv")
population_data <- read_csv("Data/ACSDP1Y2019.DP05-2024-11-18T022938.csv")

# Combine datasets into a named list
datasets <- list(states_data = states_data,
                 us_data = us_data,
                 covid_racial_data = covid_racial_data,
                 population_data = population_data)

# Check for duplicates in each dataset
for (name in names(datasets)) {
  data <- datasets[[name]]
  duplicates <- data[duplicated(data), ]  # Find duplicates
  cat(paste("Number of duplicates in", name, ":", nrow(duplicates)), "\n")
  if (nrow(duplicates) > 0) print(duplicates)
}

```

Column names in `population_data` were reformatted to follow a consistent naming convention, replacing blank spaces with underscores for easier referencing. No missing values were found.

```{r, include=FALSE}
# Extracting the overall US population estimate
population_data$Label[1] <- "US"

# Rename columns in the population_data dataset
population_data <- population_data %>%
  rename(State = `Label`,
         Population_Total = `Total population`,
         Population_Multiracial = `Two or more races`,
         Population_White = `White`,
         Population_Black = `Black or African American`,
         Population_AIAN = `American Indian and Alaska Native`,
         Population_Asian = `Asian`,
         Population_NHPI = `Native Hawaiian and Other Pacific Islander`,
         Population_Ethnicity_Hispanic = `Hispanic or Latino (of any race)`,
         Population_Ethnicity_Nonhispanic = `Not Hispanic or Latino`)


# Check for missing values
colSums(is.na(population_data))

```

The `us_states_data` dataset was created by merging `states_data` with national-level data from `us.csv` to integrate state-level and national trends. Population data (`Population_Total`) was merged into `us_states_data` to enable rate calculations for cases and deaths. The `us_states_data` dataset was filtered to retain only the states present in `population_data`, ensuring alignment between datasets. Next, we further summarized the state-level data to retain only the last recorded observation for each state within each month. To achieve this, we first extracted the `year` and `month` from the `date` column, which allowed us to group the data by `state`, `year`, and `month`. Within each group, we filtered the data to include only the row corresponding to the most recent date for that month, ensuring that the dataset captures the latest recorded statistics for each state at the monthly level.

```{r, include=FALSE}
################################################################################
#        DATA PROCESSING AND CLEANING FOR THE `states_data` & `us_data`        #
################################################################################ 

# Checking dimensions for the us_state and states_data
dim(us_data) # [1] 1158    3   
dim(states_data) # [1] 61942     5

# Merge the states_data and us_data by 'date' column 
us_states_data <- merge(states_data, us_data, by = "date", 
                        suffixes = c("_state", "_us")) %>%
  select(-fips)  # Drop the 'fips' column as it is not needed for next analysis steps

# Check for missing values
colSums(is.na(us_states_data))
# The following is the console output for missing values in population-related columns:
# 
# State                 Population_Total           Population_Multiracial 
#   0                            0                              0 
# Population_White                 Population_Black                  Population_AIAN 
#        0                                0                                0 
# Population_Asian                  Population_NHPI    Population_Ethnicity_Hispanic 
#      0                                0                                0 
# Population_Ethnicity_Nonhispanic 
#                0 

# Get the earliest and latest dates
summary(us_states_data$date)
#         Min.      1st Qu.       Median         Mean      3rd Qu.         Max. 
#   "2020-01-21"  "2020-12-09"  "2021-09-16" "2021-09-13" "2022-06-20" "2023-03-23" 


# Filter `us_states_data` to include only states present in `population_data$Label`
filtered_us_states_data1 <- us_states_data %>%
  filter(state %in% unique(population_data$State))


# Merge only the Population_Total column into filtered_us_states_data
filtered_us_states_data1 <- filtered_us_states_data1 %>%
  left_join(population_data %>% 
              select(State, Population_Total), 
    by = c("state" = "State"))  # Match on state 

# Add rate columns for cases and deaths
filtered_us_states_data1 <- filtered_us_states_data1 %>%
  mutate(state_cases_rate = (cases_state / Population_Total) * 100000,   # Cases rate per 100,000
         state_deaths_rate = (deaths_state / Population_Total) * 100000) # Deaths rate per 100,000
  

# Extract the total U.S. population from the first row of Population_Total in population_data
us_population <- population_data$Population_Total[1]

# Add rate columns for cases and deaths at the national level
filtered_us_states_data1 <- filtered_us_states_data1 %>%
  mutate(us_cases_rate = (cases_us / us_population) * 100000,    # Cases rate per 100,000 for the US
         us_deaths_rate = (deaths_us / us_population) * 100000)  # Deaths rate per 100,000 for the US


# Filter to retain the last recorded date per month for each state
monthly_data <- filtered_us_states_data1 %>%
  mutate(year = year(date),                      # Extract the year from the date    
         month = month(date, label = TRUE)) %>%  # Extract the month as a label
  group_by(state, year, month) %>%               # Group by state, year, and month
  filter(date == max(date)) %>%                  # Keep rows where the date is the maximum within each group
  ungroup() %>%
  select(date, month, year, everything())        # Rearrange columns: date, month, year, and all others

  
```

Below is the summary information about the variables in the cleaned `monthly_data` dataset:

The **date** variable represents the latest recorded observation date for each month and state, with additional columns for **month** and **year** extracted from the date. The **state** variable identifies the location where the data was collected. Key measures include **cases_state** and **deaths_state**, representing the cumulative COVID-19 cases and deaths reported within each state by the recorded date. National-level statistics are provided through **cases_us** and **deaths_us**, which reflect cumulative COVID-19 cases and deaths across the United States. Population data is captured in the **Population_Total** variable, sourced from the American Community Survey estimates. **state_cases_rate** and **state_deaths_rate** provide state-level case and death rates per 100,000 population, while **us_cases_rate** and **us_deaths_rate** offer the same metrics at the national level.

### Processing and Cleaning the COVID Racial Data

We also processed and cleaned the COVID racial data to ensure its suitability for analysis and to address issues such as missing values and irrelevant observations. The `Date` column was converted to a standard date format to enable temporal analysis. A summary of the dataset revealed a date range from April 12, 2020, to March 7, 2021, and a total of 5,320 rows with 54 variables. We assessed missing values across variables and states.

```{r, include=FALSE}
################################################################################
#        DATA PROCESSING AND CLEANING FOR THE `covid_racial_data`              #
################################################################################ 

# COVID Racial Data processing and cleaning

# Convert the column to Date format
covid_racial_data$Date <- as.Date(as.character(covid_racial_data$Date), format = "%Y%m%d")

# Get the earliest and latest dates
summary(covid_racial_data$Date)
#    Min.      1st Qu.       Median         Mean      3rd Qu.         Max. 
# "2020-04-12" "2020-07-01" "2020-09-23" "2020-09-23" "2020-12-16" "2021-03-07" 

dim(covid_racial_data) # [1] 5320   54

names(covid_racial_data)

```

A state-wise analysis identified the top 10 states/territories with the highest overall missing values, helping to pinpoint data quality issues specific to geographic regions. These included; American Samoa, Northern Mariana Islands, Puerto Rico, Virgin Islands, North Dakota, Guam, West Virginia, Hawaii, New York, and Pennsylvania. Column names were cleaned for consistency, replacing spaces with underscores and standardizing terms like "Hosp" to "Hospitalized." Racial categories such as "AIAN" and "NHPI" were kept in uppercase for clarity.

```{r, include=FALSE}
# 1. Missingness by state
missing_by_state <- covid_racial_data %>%
  group_by(State) %>%
  summarise(across(everything(), ~ sum(is.na(.)), .names = "Missing_{col}")) %>%
  mutate(Total_Missing = rowSums(select(., starts_with("Missing_")))) %>%
  arrange(desc(Total_Missing))  # Sort by total missing values in descending order

# Extract the top 10 state names
top_10_states <- missing_by_state %>%
  slice_head(n = 10) %>%  # Select the first 10 rows (sorted by Total_Missing)
  pull(State)  # Extract the 'State' column as a vector

# View the top 10 state names
top_10_states

# Custom function to clean column names
format_column_names <- function(names) {
  names %>%
    str_replace_all("_", " ") %>%  # Replace underscores with spaces
    str_replace("Hosp", "Hospitalized") %>%  # Replace "Hosp" with "Hospitalized"
    str_to_title() %>%  # Convert to title case
    str_replace_all("\\b(Aian|Nhpi)\\b", toupper)  # Keep AIAN and NHPI in uppercase
}

# Apply the custom function to reformat variable names
covid_racial_data <- covid_racial_data %>%
  rename_with(format_column_names)

```

A column-wise analysis quantified the percentage of missing data for each variable, identifying variables with significant gaps as visualized in Figure.

```{r, fig.height=9,fig.width=8}
# 3. Visualize missingness using naniar
# Matrix plot
gg_miss_var(covid_racial_data) +
  labs(title = "Missingness in the Covid Racial Data", x = "Variables", y = "Number of Missing Values") +
  theme_minimal() 

```

For the next step of the analysis, we chose to exclude the following territories (Puerto Rico, Virgin Islands, Guam, Northern Mariana Islands, American Samoa) because they had high levels of missingness and their population estimates were unavailable in the `population_data`.

```{r, include=FALSE}
# Replace blank spaces in column names with underscores
names(covid_racial_data) <- gsub(" ", "_", names(covid_racial_data))

# Define the territories to exclude due to high missingness and lack of population estimates
territories_to_exclude <- c("VI", "AS", "GU", "MP", "PR")

# Filter out the specified territories
filtered_racial_data <- covid_racial_data %>%
  filter(!State %in% territories_to_exclude) %>%
  select(c(1:2, 4:5, 7:10, 13:14, 17:18, 20:23, 26:27, 30:31, 33:36, 39:40, 43:44, 46:49, 52:53))

dim(filtered_racial_data) # [1] 4845   34
```

Since the data collected was cumulative, missing values for key metrics (e.g., cases, deaths, hospitalizations, tests) were forward-filled within each state group. This ensured that subsequent analyses reflected the most recent available data, even for partially missing time points. Finally, the `monthly_racial_data` dataset was aggregated to retain only the last recorded observation for each month, grouped by state. This step reduced temporal granularity while maintaining monthly summaries essential for trend analysis.

```{r}
# Replace missing observations with the last available value
filtered_racial_data <- filtered_racial_data %>%
  group_by(State) %>% 
  arrange(Date) %>%  # Sort dates in ascending order
  fill(Cases_White:Tests_Ethnicity_Hispanic, .direction = "down") %>%  # Fill missing values forward
  ungroup()


# Filter to retain the last recorded date per month for each state
monthly_racial_data <- filtered_racial_data %>%
  mutate(year = year(Date),                      # Extract the year from the date    
         month = month(Date, label = TRUE)) %>%  # Extract the month as a label
  group_by(State, year, month) %>%               # Group by state, year, and month
  filter(Date == max(Date)) %>%                  # Keep rows where the date is the maximum within each group
  ungroup() %>%
  select(month, year, everything(), -Date)       # Rearrange columns: date, month, year, and all others

```

The cleaned `monthly_racial_data` dataset provides detailed demographic information for analyzing COVID-19 cases, deaths, hospitalizations, and testing across racial and ethnic groups at the state level. The **month** and **year** variables indicate when the observations were recorded, while **State** represents the state using its two-letter abbreviation. The dataset includes cumulative counts of COVID-19 cases, deaths, hospitalizations, and tests, disaggregated by racial and ethnic groups. For racial categories, it provides variables for **White**, **Black**, **Asian**, **American Indian or Alaska Native (AIAN)**, **Native Hawaiian or Pacific Islander (NHPI)**, and **Multiracial** populations. Ethnicity-specific variables capture data for **Hispanic** individuals and **non-Hispanic** individuals, with further distinctions between cases, deaths, hospitalizations, and tests.

### Aggregation of State-Level, COVID Racial, and Population Data

To prepare the combined state-level COVID-19 racial dataset for analysis, first,the `filtered_us_states_data` used to create the `monthly_data` in step was filtered to include only data between April 12, 2020, and March 7, 2021. This date range corresponds to the period covered by the `covid_racial_data`. Similarly, regions such as Puerto Rico, Virgin Islands, Guam, Northern Mariana Islands, and American Samoa were excluded due to significant missing data and the lack of reliable population estimates. This step allowed the focus to remain on states with complete and consistent data. Next, the `monthly_us_state_data` was created by only retaining the latest available observation (based on date) for each state, year and month. State names in the `monthly_us_state_data` and `population_data` datasets were converted to standardized abbreviations. This ensured compatibility when merging datasets and avoided mismatches due to variations in naming conventions.

```{r, include=FALSE}
# Define the date range
start_date <- as.Date("2020-04-12")
end_date <- as.Date("2021-03-07")

# Filter the merged dataset by the date range
filtered_us_states_data_2 <- filtered_us_states_data1 %>%
  filter(date >= start_date & date <= end_date) %>% # 17160 by 11
  filter(!state %in% c("Puerto Rico", "Virgin Islands", "Guam", "Northern Mariana Islands", "American Samoa")) %>%
  arrange(date)

dim(filtered_us_states_data_2) # [1] 16830    11

# Filter to retain the last recorded date per month for each state
monthly_us_state_data <- filtered_us_states_data_2 %>%
  rename(State = state) %>%
  mutate(year = year(date),                      # Extract the year from the date  
         month = month(date, label = TRUE)) %>%  # Extract the month as a label
  group_by(State, year, month) %>%               # Group by state, year, and month
  filter(date == max(date)) %>%                  # Keep rows where the date is the maximum within each group
  ungroup() %>%
  select(month, year, everything(), -date)       # Include all columns but exclude the `date` column


# Convert full state names in monthly_us_state_data to state abbreviations
monthly_us_state_data$State<- state2abbr(monthly_us_state_data$State)

# Convert full state names in population_data to state abbreviations
population_data$State<- state2abbr(population_data$State)

```

Secondly, the `monthly_us_state_data` dataset was joined with the `monthly_racial_data` and `population_data`, to create the `combined_monthly_data` that has the integration of COVID-19 racial statistics with population metrics.

```{r}
# Join the datasets
combined_monthly_data <- monthly_us_state_data %>%
  left_join(monthly_racial_data, by = c("month", "year", "State")) %>%
  left_join(population_data, by = c("State", "Population_Total"))

```

Thirdly, in the `combined_monthly_data`, Rates for cases, deaths, hospitalizations, and tests were calculated for each racial/ethnic group by dividing the respective counts by the population of that group and scaling to 100,000. These standardized rates facilitate meaningful comparisons across states and demographic groups.

```{r}
# Calculate rates for columns 13 to 44 based on population values from columns 45 to 52
combined_monthly_data <- combined_monthly_data %>%
  mutate(
    across(
      .cols = 13:44,  # Columns for cases, deaths, hospitalizations, tests
      .fns = ~ . / get(sub("Cases|Deaths|Hospitalized|Tests", "Population", cur_column())) * 100000,
      .names = "{.col}_rate"
    )
  )

```

The cleaned dataset, `combined_monthly_data`, includes variables for monthly cases, deaths, hospitalizations, and tests for various racial and ethnic groups, along with population data and calculated rates. Below is a preview of the dataset structure:
:::

::: {#refs}
:::
