---
title: "Racial Disparities in COVID-19 Outcomes"
subtitle: "<div style='text-align: center; font-size: 1.2rem;'>Rophence Ojiambo<br>December 16, 2024</div>"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    theme: materia
    highlight-style: github
    page-layout: full
    title-block-align: center
    self_contained: false
    toc: false
    include-in-header:
      text: |
        <style>
          h1 { text-align: center; font-size: 3rem; }
          .subtitle { text-align: center; margin-top: -10px; margin-bottom: 20px; }
        </style>
lightbox: true
link-citations: true
colorlinks: true
linkcolor: blue
urlcolor: blue
bibliography: references.bib
csl: apa-numeric-superscript-brackets.csl
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      dpi = 300)
```

COVID-19 had an unprecedented impact on public health, particularly among vulnerable populations. This project examines the disproportionate effects of the pandemic on racial and ethnic communities in the United States. 

Below, you can navigate through the project components to explore the methodology, analysis, and findings.


::: {.panel-tabset .tabset-scrollable}

# Introduction

## Overview

The novel Coronavirus disease (COVID-19), caused by the serious acute respiratory syndrome (SARS-CoV-2) [@synowiec2021severe] virus is a highly infectious respiratory disease, first identified in Wuhan, China, in December 2019 [@zhu2020novel]. Since its emergence, the pandemic has impacted millions worldwide, with over 770 million confirmed cases and approximately 7 million deaths reported globally by the [**World Health Organization’s (WHO) Coronavirus (COVID-19) Dashboard**](https://data.who.int/dashboards/covid19/cases?n=c). The COVID-19 pandemic, which was first traced in the United States in January 2020 [@holshue2020first], highlighted and worsened existing health disparities, with racial and ethnic minority groups being disproportionately affected. Communities such as Black, Hispanic, Asian, and Native American populations faced higher rates of infection [@hooper2020covid], hospitalizations [@garg2020hospitalization; @romano2021trends], and mortality compared to White populations [@rossen2020excess; @gold2020race; @tai2021disproportionate]. Understanding the patterns and magnitude of these disparities is critical for informing targeted public health interventions for future pandemics.

## Problem Statement

This project aims to investigate the disproportionate impact of the COVID-19 pandemic on racial and ethnic communities in the United States. By analyzing disparities in cases, deaths, hospitalizations, and testing rates across states and racial/ethnic groups, the research seeks to illuminate the differential burden of COVID-19 on these communities. The insights derived from this study will be vital for public health planning and policy making to promote equity and preparedness for future public health emergencies.

## Data and Methodology

To explore these disparities, I will utilize the following datasets:

-   **Nation level: us.csv**
-   **State level: us-states.csv**

The above two datasets contain the US COVID-19 data (cumulative cases and deaths) till 03/23/2023, and were originally obtained from [The New York Times](https://www.nytimes.com/interactive/2021/us/covid-cases.html), [Github repository](https://github.com/nytimes/covid-19-data)

Additionally, I will use the following external data sets to improve the analysis on the required data:

-   [**COVID Racial Data Tracker**](https://covidtracking.com/race): Provides state-level information on cumulative COVID-19 cases and deaths (overall and by racial/ethnic group), as well as the number of hospitalizations and tests conducted by racial/ethnic group.
-   [**American Community Survey (ACS) Demographic and Housing 5-Year Estimates (2019)**](https://data.census.gov/table/ACSDP5Y2019.DP05): Contains U.S. population and state-level population estimate data for 2019, enabling the calculation of rates for cases, deaths, hospitalizations, and testing across racial and ethnic groups.

These datasets will be merged to facilitate a comprehensive analysis of state-level disparities.

## Proposed Analytic Approach

The analysis will involve the following steps:

1.  **Exploratory Data Analysis:**

-   Data cleaning, transformation of variables, missing data,

2.  **Descriptive Analysis and Data Visualization:**

-   Overall description of Total Covid Cases and Deaths
-   Calculate national and state-level rates for COVID-19 cases, deaths, hospitalizations, and tests.
-   Use visualizations such as bar charts and maps to compare these counts/rates across states and changes over time.

3.  **Correlation Analysis:**

-   COVID-19 cases, deaths, hospitalizations, and tests by racial and ethnic groups and changes over time as the pandemic progressed.
-   Fit a multiple linear regression model to examine the relationship between race, number of tests, number of hospitalizations and COVID-19 deaths.

## Contribution

This analysis will shed light on how the COVID-19 pandemic disproportionately affected racial and ethnic minorities in the United States. By examining and comparing the differences in cases, deaths, and hospitalizations across populations and states, this project will highlight areas where inequities are most severe. The findings can help leaders and public health officials focus their efforts, distribute resources fairly, and address the underlying issues that make some communities more vulnerable during health crises.

<div style="text-align: center; margin-top: 20px;">
  <a href="#top" style="text-decoration: none; background-color: #007BFF; color: white; padding: 10px 20px; border-radius: 5px; font-size: 16px;">Back to Top</a>
</div>


# Packages Required

Next, we load all the packages required for the analysis

```{r}
# Loading required libraries

library(readr)       # For reading CSV and other data files efficiently into R.
library(dplyr)       # For data manipulation: filtering, summarizing, joining, and transforming datasets.
library(tidyr)       # For reshaping and cleaning data: pivoting, filling missing values.
library(kableExtra)  # For creating and customizing tables in RMarkdown and reports.
library(DT)          # For creating interactive tables in HTML reports or Shiny apps.

library(lubridate)   # For working with dates and times: parsing, extracting, and manipulating date components.
library(stringr)     # For handling string operations: pattern matching, substitution, and string manipulation.

library(naniar)      # For identifying, visualizing, and handling missing data in datasets.
library(usdata)      # For working with U.S.-specific datasets, such as state and county names, populations, and abbreviations.

library(ggplot2)     # For creating data visualizations.
library(gridExtra)   # For combining and arranging multiple plots into a single layout.
library(plotly)      # For creating interactive and dynamic visualizations, such as zooming and hovering features.
library(scales)      # For formatting numeric scales: adding commas, percentages, or scientific notation.
library(broom)       # Tidying output of statistical models

```



# Data Preparation



## Original Source

### National-level and State-level data

The national-level (`us.csv`) and state-level (`us-states.csv`) COVID-19 datasets were sourced from [The New York Times](https://www.nytimes.com/interactive/2021/us/covid-cases.html) and are publicly accessible through their [GitHub repository](https://github.com/nytimes/covid-19-data). These datasets document cumulative COVID-19 cases and deaths across the United States up to March 23, 2023. Compiled by a dedicated team of journalists, the data was gathered by monitoring press conferences, analyzing public health reports, and seeking clarification from officials to ensure accuracy and completeness. This effort addressed challenges in a fragmented public health system, where overwhelmed state, county, and territorial agencies often struggled with consistent and timely reporting. The datasets include both confirmed and probable cases and deaths, adhering to definitions established by the Council of State and Territorial Epidemiologists and the Centers for Disease Control and Prevention (CDC). Corrections or reassignments between counties or states sometimes led to declines in reported cases or deaths on certain dates. The case counts represent all cases, including individuals who recovered or died. Updated daily, the datasets reflect real-time changes, with retroactive corrections applied when errors were identified or new information became available. To preserve transparency, the datasets retain their raw reported values with no imputed values.

#### Variables Overview

1. **`us.csv` (National-Level Data)**: has 1,158 observations and 3 variables (`date`, `cases`, `deaths`)
2. **`us-states.csv` (State-Level Data)**: has 61,942 observations and 5 variables (`date`, `state`, `fips` (Federal Information Processing Standards (FIPS) code for geographic identification), `cases`, `deaths`)



### American Community Survey (ACS) Demographic and Housing 5-Year Estimates (2019)


The data for this analysis was obtained from the [U.S. Census Bureau's American Community Survey (ACS) 5-Year Estimates](https://data.census.gov/table/ACSDP5Y2019.DP05?t=Populations%20and%20People&g=010XX00US_040XX00US01,02,04,05,06,08,09,10,11,12,13,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35,36,37,38,39,40,41,42,44,45,46,47,48,49,50,51,53,54,55,56,60,66,69,72,78). The Census Bureau's [Population Estimates Program (PEP)](https://www.census.gov/programs-surveys/popest/about.html) produces estimates of the population for the United States, its states, counties, cities, and towns, as well as for Puerto Rico. PEP annually utilizes current data on births, deaths, and migration to calculate population change since the most recent decennial census and produce a time series of estimates of population, demographic components of change, and housing units. These estimates are used in federal funding allocations, as survey controls, as denominators for vital rates and per capita time series, and as indicators of recent demographic changes. The ACS 5-Year Estimates (2019) include data collected from January 1, 2015, to December 31, 2019. As a sample survey, the ACS provides estimates accompanied by margins of error (MOE).

#### Variables Overview


- **Label**: Represents the name of the geographic region (e.g., United States, individual states).
- **Total population**: The total population estimate for each geographic region.
- **Two or more races**: Population estimate for individuals who identify with two or more racial groups.
- **White**: Population estimate for individuals identifying as White.
- **Black or African American**: Population estimate for individuals identifying as Black or African American.
- **American Indian and Alaska Native**: Represents individuals identifying as Native American or Alaska Native.
- **Asian**: Population estimate for individuals identifying as Asian.
- **Native Hawaiian and Other Pacific Islander**: Represents individuals identifying as Native Hawaiian or Pacific Islander.
- **Hispanic or Latino (of any race)**: Population estimate for individuals identifying as Hispanic or Latino, regardless of their racial category.
- **Not Hispanic or Latino**: Population estimate for individuals identifying as non-Hispanic.


### COVID Racial Data Tracker data

The CRDT (COVID Racial Data Tracker) data, labelled `CRDT Data - CRDT.csv` was originally obtained from the [COVID Racial Data Tracker](https://covidtracking.com/race/about), a project developed in partnership with The [Atlantic’s COVID Tracking Project](https://covidtracking.com/) and the [Boston University Center for Antiracist Research](https://www.bu.edu/antiracism-center/), founded by Activist and scholar Dr. Ibram X. Kendi. This dataset provides information on the racial and ethnic disparities in COVID-19 cases, deaths, hospitalizations, and tests conducted across 51 states/territories, meaning granular insights at the county or local levels are not directly available.. The data is publicly available and was accessed from the COVID Racial Data Tracker website, which served as a key resource during the pandemic to highlight inequities in COVID-19 outcomes. The original dataset, which can be directly using this [link](https://docs.google.com/spreadsheets/d/e/2PACX-1vS8SzaERcKJOD_EzrtCDK1dX1zkoMochlA9iHoHg_RSw3V8bkpfk1mpw4pfL5RdtSOyx_oScsUtyXyk/pub?gid=43720681&single=true&output=csv), includes **5,320 rows** and **54 variables**, covering data collected from **April 12, 2020**, to **March 7, 2021**, across various states/territories in the United States. The Racial Data Tracker aimed to collect racial and demographic data to understand the outbreak and protect vulnerable communities. In the original dataset, missing values were represented as blank cells. Upon loading the data into R, these blanks were automatically interpreted as `NA`. These missing entries occur because not all states reported COVID-19 data disaggregated by race or ethnicity consistently during the pandemic.


#### Variables Overview

1.  **Date and State**:
    -   `Date`: The reporting date in the format `YYYY-MM-DD`.
    -   `State`: The U.S. state or territory where data was reported.
2.  **COVID-19 Outcomes**:
    -   **Cases**: Total and race/ethnicity-specific cumulative COVID-19 cases (e.g., `Cases_Total`, `Cases_White`, `Cases_Black`).
    -   **Deaths**: Total and race/ethnicity-specific cumulative deaths due to COVID-19 (e.g., `Deaths_Total`, `Deaths_White`, `Deaths_Black`).
    -   **Hospitalizations (`Hosp_`)**: Total and race/ethnicity-specific cumulative hospitalizations for COVID-19 (e.g., `Hosp_Total`, `Hosp_Black`).
    -   **Tests (`Tests_`)**: Total and race/ethnicity-specific COVID-19 tests conducted (e.g., `Tests_Total`, `Tests_Latinx`).

##### Racial and Ethnic Groups

The dataset categorizes individuals into the following racial and ethnic groups:

: Racial and Ethnic Groups in the Dataset {#tbl-racial-ethnic-groups}

| **Variable**             | **Description**                                                                 |
|--------------------------|---------------------------------------------------------------------------------|
| **White**                | Individuals identified as non-Hispanic White.                                  |
| **Black**                | Individuals identified as non-Hispanic Black or African American.              |
| **Latinx**               | Individuals identified as Hispanic or Latino.                                  |
| **Asian**                | Individuals identified as non-Hispanic Asian.                                  |
| **AIAN**                 | Individuals identified as non-Hispanic American Indian or Alaska Native.       |
| **NHPI**                 | Individuals identified as non-Hispanic Native Hawaiian or Pacific Islander.    |
| **Multiracial**          | Individuals identified as belonging to multiple racial groups.                 |
| **Other**                | Individuals reported in categories not otherwise specified.                    |
| **Unknown**              | Individuals whose racial or ethnic data were not reported or categorized. Missing values were left blank and interpreted as `NA` in R. |
| **Hispanic Ethnicity**   | Individuals identified as Hispanic Ethnicity                                   |
| **Non-Hispanic Ethnicity** | Individuals identified as Non-Hispanic Ethnicity.                            |
| **Ethnicity Unknown**    | Individuals whose ethnicity was not recorded.                                  |




## Data Importing and Cleaning

### Processing and Cleaning the state-level and national-level, and population estimates Data sets

Four datasets were loaded for analysis: `us-states.csv` (state-level COVID-19 data), `us.csv` (national-level COVID-19 data), `CRDT Data - CRDT.csv` (COVID Racial Data Tracker), and `ACSDP1Y2019.DP05-2024-11-18T022938.csv` (American Community Survey population data). Each dataset was checked for duplicate rows using the `duplicated()` function, and no duplicates were found in any of the datasets. 


```{r, results='hide'}
################################################################################
#                  DATA IMPORTING & CHECKING FOR DUPLICATES                    #
################################################################################ 

# Load the datasets

states_data <- read_csv("Data/us-states.csv")
us_data <- read_csv("Data/us.csv")
covid_racial_data <- read_csv("Data/CRDT Data - CRDT.csv")
population_data <- read_csv("Data/ACSDP1Y2019.DP05-2024-11-18T022938.csv")

# Combine datasets into a named list
datasets <- list(states_data = states_data,
                 us_data = us_data,
                 covid_racial_data = covid_racial_data,
                 population_data = population_data)

# Check for duplicates in each dataset
for (name in names(datasets)) {
  data <- datasets[[name]]
  duplicates <- data[duplicated(data), ]  # Find duplicates
  cat(paste("Number of duplicates in", name, ":", nrow(duplicates)), "\n")
  if (nrow(duplicates) > 0) print(duplicates)
}

```


Column names in `population_data` were reformatted to follow a consistent naming convention, replacing blank spaces with underscores for easier referencing. No missing values were found.


```{r, results='hide'}
# Extracting the overall US population estimate
population_data$Label[1] <- "US"

# Rename columns in the population_data dataset
population_data <- population_data %>%
  rename(State = `Label`,
         Population_Total = `Total population`,
         Population_Multiracial = `Two or more races`,
         Population_White = `White`,
         Population_Black = `Black or African American`,
         Population_AIAN = `American Indian and Alaska Native`,
         Population_Asian = `Asian`,
         Population_NHPI = `Native Hawaiian and Other Pacific Islander`,
         Population_Ethnicity_Hispanic = `Hispanic or Latino (of any race)`,
         Population_Ethnicity_Nonhispanic = `Not Hispanic or Latino`)


# Check for missing values
colSums(is.na(population_data))

```

Population data (`Population_Total`) was merged into `states_data` to enable rate calculations for cases and deaths. The `states_data` dataset was filtered to retain only the states present in `population_data`, ensuring alignment between datasets. Next, we further summarized the state-level data to retain only the last recorded observation for each state within each month. To achieve this, we first extracted the `year` and `month` from the `date` column, which allowed us to group the data by `state`, `year`, and `month`. Within each group, we filtered the data to include only the row corresponding to the most recent date for that month, ensuring that the dataset captures the latest recorded statistics for each state at the monthly level. 




```{r, include=FALSE}
################################################################################
#        DATA PROCESSING AND CLEANING FOR THE `states_data` & `us_data`        #
################################################################################ 

# Checking dimensions for the us_state and states_data
dim(us_data)     # [1] 1158    3   
dim(states_data) # [1] 61942     5

states_data <- states_data %>%
  select(-fips)  # Drop the 'fips' column as it is not needed for next analysis steps

# Check for missing values
colSums(is.na(states_data))
# The following is the console output for missing values in population-related columns:
#   date  state  cases deaths 
#     0      0      0      0 

# Get the earliest and latest dates
summary(states_data$date)
#         Min.      1st Qu.       Median         Mean      3rd Qu.         Max. 
#   "2020-01-21"  "2020-12-09"  "2021-09-16" "2021-09-13" "2022-06-20" "2023-03-23" 


# Filter `states_data` to include only states present in `population_data$Label`
filtered_states_data <- states_data %>%
  filter(state %in% unique(population_data$State))


# Merge only the Population_Total column into filtered_states_data
filtered_states_data <- filtered_states_data %>%
  left_join(population_data %>% 
              select(State, Population_Total), 
    by = c("state" = "State"))  # Match on state 

filtered_states_data <- filtered_states_data %>%
  mutate(state_cases_rate = round((cases/ Population_Total) * 100000, 8),   # Cases rate per 100,000
         state_deaths_rate = round((deaths / Population_Total) * 100000, 8)) # Deaths rate per 100,000

#------------------------------------------------------------------------------#

# Extract the total U.S. population from the first row of Population_Total in population_data
us_population <- population_data$Population_Total[1]


# Add rate columns for cases and deaths at the national level
us_data <- us_data %>%
  mutate(us_cases_rate = round((cases / us_population) * 100000, 8),    # Cases rate per 100,000 for the US
         us_deaths_rate = round((deaths / us_population) * 100000, 8))  # Deaths rate per 100,000 for the US



# Filter to retain the last recorded date per month for each state
monthly_state_data <- filtered_states_data %>%
  mutate(year = year(date),                      # Extract the year from the date    
         month = month(date, label = TRUE)) %>%  # Extract the month as a label
  group_by(state, year, month) %>%               # Group by state, year, and month
  filter(date == max(date)) %>%                  # Keep rows where the date is the maximum within each group
  ungroup() %>%
  select(date, month, year, everything())        # Rearrange columns: date, month, year, and all others

  
```

Below is the summary information about the variables in the cleaned `monthly_state_data` and `monthly_us_data` datasets:

The **date** variable represents the latest recorded observation date for each month and state, with additional columns for **month** and **year** extracted from the date. The **state** variable identifies the location where the data was collected. Key measures include **cases_state** and **deaths_state**, representing the cumulative COVID-19 cases and deaths reported within each state by the recorded date. National-level statistics are provided through **cases_us** and **deaths_us**, which reflect cumulative COVID-19 cases and deaths across the United States. Population data is captured in the **Population_Total** variable, sourced from the American Community Survey estimates. **state_cases_rate** and **state_deaths_rate** provide state-level case and death rates per 100,000 population, while **us_cases_rate** and **us_deaths_rate** offer the same metrics at the national level. 


Below is a preview of the cleaned `monthly_state_data` dataset:

```{r}
# Preview of the cleaned monthly state data
datatable(monthly_state_data, colnames = c("Date", "Month", "Year", "State", "State-Level Cases", "State-Level Deaths", 
                                     "Total Population", "State Case Rate per 100k", 
                                     "State Death Rate per 100k"),
          filter = "top")
```

Similarly, we filtered the `us_data` to retain monthly observations since the data was cumulative.Below is a preview of the cleaned `monthly_us_data` dataset:

```{r}
# Filter to retain the last recorded date per month for the national data
monthly_us_data <- us_data %>%
  mutate(year = year(date),                      # Extract the year from the date    
         month = month(date, label = TRUE)) %>%  # Extract the month as a label
  group_by(year, month) %>%                      # Group by year, and month
  filter(date == max(date)) %>%                  # Keep rows where the date is the maximum within each group
  ungroup() %>%
  select(date, month, year, everything())        # Rearrange columns: date, month, year, and all others

# Preview of the cleaned monthly national data
datatable(monthly_us_data, colnames = c("Date", "Month", "Year", "US Total Cases", "US Total Deaths",
                                           "National Case Rate per 100k", "National Deaths Rate per 100k"),
          filter = "top")

```



### Processing and Cleaning the COVID Racial Data

We also processed and cleaned the COVID racial data to ensure its suitability for analysis and to address issues such as missing values and irrelevant observations. The `Date` column was converted to a standard date format to enable temporal analysis. A summary of the dataset revealed a date range from April 12, 2020, to March 7, 2021, and a total of 5,320 rows with 54 variables. We assessed missing values across variables and states. 


```{r, results='hide'}
################################################################################
#        DATA PROCESSING AND CLEANING FOR THE `covid_racial_data`              #
################################################################################ 

# COVID Racial Data processing and cleaning

# Convert the column to Date format
covid_racial_data$Date <- as.Date(as.character(covid_racial_data$Date), format = "%Y%m%d")

# Get the earliest and latest dates
summary(covid_racial_data$Date)
#    Min.      1st Qu.       Median         Mean      3rd Qu.         Max. 
# "2020-04-12" "2020-07-01" "2020-09-23" "2020-09-23" "2020-12-16" "2021-03-07" 

dim(covid_racial_data) # [1] 5320   54

names(covid_racial_data)

```



A state-wise analysis identified the top 10 states/territories with the highest overall missing values, helping to pinpoint data quality issues specific to geographic regions. These included; American Samoa, Northern Mariana Islands, Puerto Rico, Virgin Islands, North Dakota, Guam, West Virginia, Hawaii, New York, and Pennsylvania. Column names were cleaned for consistency, replacing spaces with underscores and standardizing terms like "Hosp" to "Hospitalized." Racial categories such as "AIAN" and "NHPI" were kept in uppercase for clarity.

```{r, results='hide'}
# Missingness by state
missing_by_state <- covid_racial_data %>%
  group_by(State) %>%
  summarise(across(everything(), ~ sum(is.na(.)), .names = "Missing_{col}")) %>%
  mutate(Total_Missing = rowSums(select(., starts_with("Missing_")))) %>%
  arrange(desc(Total_Missing))  # Sort by total missing values in descending order

# Extract the top 10 state names
top_10_states <- missing_by_state %>%
  slice_head(n = 10) %>%  # Select the first 10 rows (sorted by Total_Missing)
  pull(State)  # Extract the 'State' column as a vector

# View the top 10 state names
top_10_states
# [1] "AS" "MP" "PR" "VI" "ND" "GU" "WV" "HI" "NY" "PA"

# Custom function to clean column names
format_column_names <- function(names) {
  names %>%
    str_replace_all("_", " ") %>%  # Replace underscores with spaces
    str_replace("Hosp", "Hospitalized") %>%  # Replace "Hosp" with "Hospitalized"
    str_to_title() %>%  # Convert to title case
    str_replace_all("\\b(Aian|Nhpi)\\b", toupper)  # Keep AIAN and NHPI in uppercase
}

# Apply the custom function to reformat variable names
covid_racial_data <- covid_racial_data %>%
  rename_with(format_column_names)

```


A column-wise analysis quantified the percentage of missing data for each variable, identifying variables with significant gaps as visualized in Figure. 



```{r, fig.height=9,fig.width=8}
# Visualize missingness using naniar

gg_miss_var(covid_racial_data) +
  labs(title = "Missingness in the Covid Racial Data", x = "Variables", y = "Number of Missing Values") +
  theme_minimal() 
```


For the next step of the analysis, we chose to exclude the following territories (Puerto Rico, Virgin Islands, Guam, Northern Mariana Islands, American Samoa) because they had high levels of missingness and their population estimates were  unavailable in the `population_data`. 

 

```{r, results='hide'}
# Replace blank spaces in column names with underscores
names(covid_racial_data) <- gsub(" ", "_", names(covid_racial_data))

# Define the territories to exclude due to high missingness and lack of population estimates
territories_to_exclude <- c("VI", "AS", "GU", "MP", "PR")

# Filter out the specified territories
filtered_racial_data <- covid_racial_data %>%
  filter(!State %in% territories_to_exclude) %>%
  select(c(1:2, 4:5, 7:10, 13:14, 17:18, 20:23, 26:27, 30:31, 33:36, 39:40, 43:44, 46:49, 52:53))

dim(filtered_racial_data) # [1] 4845   34
```



Since the data collected was cumulative, missing values for key metrics (e.g., cases, deaths, hospitalizations, tests) were forward-filled within each state group. This ensured that subsequent analyses reflected the most recent available data, even for partially missing time points. Finally, the `monthly_racial_data` dataset was aggregated to retain only the last recorded observation for each month, grouped by state. This step reduced temporal granularity while maintaining monthly summaries essential for trend analysis. 


```{r}
# Replace missing observations with the last available value
filtered_racial_data <- filtered_racial_data %>%
  group_by(State) %>% 
  arrange(Date) %>%  # Sort dates in ascending order
  fill(Cases_White:Tests_Ethnicity_Hispanic, .direction = "down") %>%  # Fill missing values forward
  ungroup()


# Filter to retain the last recorded date per month for each state
monthly_racial_data <- filtered_racial_data %>%
  mutate(year = year(Date),                      # Extract the year from the date    
         month = month(Date, label = TRUE)) %>%  # Extract the month as a label
  group_by(State, year, month) %>%               # Group by state, year, and month
  filter(Date == max(Date)) %>%                  # Keep rows where the date is the maximum within each group
  ungroup() %>%
  select(Date, month, year, everything())       # Rearrange columns: date, month, year, and all others

```


The cleaned `monthly_racial_data` dataset provides detailed demographic information for analyzing COVID-19 cases, deaths, hospitalizations, and testing across racial and ethnic groups at the state level. The **month** and **year** variables indicate when the observations were recorded, while **State** represents the state using its two-letter abbreviation. The dataset includes cumulative counts of COVID-19 cases, deaths, hospitalizations, and tests, disaggregated by racial and ethnic groups. For racial categories, it provides variables for **White**, **Black**, **Asian**, **American Indian or Alaska Native (AIAN)**, **Native Hawaiian or Pacific Islander (NHPI)**, and **Multiracial** populations. Ethnicity-specific variables capture data for **Hispanic** individuals and **non-Hispanic** individuals, with further distinctions between cases, deaths, hospitalizations, and tests.

Below is a preview of the cleaned `monthly_racial_data` dataset:

```{r}
# Display data table
datatable(monthly_racial_data, filter = "top")
```

### Aggregation of State-Level, COVID Racial, and Population Data

To prepare the combined state-level COVID-19 racial dataset for analysis, first,the `filtered_states_data` used to create the `monthly_data` in step was filtered to include only data between April 12, 2020, and March 7, 2021. This date range corresponds to the period covered by the `covid_racial_data`. Similarly, regions such as Puerto Rico, Virgin Islands, Guam, Northern Mariana Islands, and American Samoa were excluded due to significant missing data and the lack of reliable population estimates. This step allowed the focus to remain on states with complete and consistent data. Next, the `monthly_us_state_data` was created by only retaining the latest available observation (based on date) for each state, year and month. State names in the `monthly_us_state_data` and `population_data` datasets were converted to standardized abbreviations. This ensured compatibility when merging datasets and avoided mismatches due to variations in naming conventions.


```{r, results='hide'}
# Define the date range
start_date <- as.Date("2020-04-12")
end_date <- as.Date("2021-03-07")

# Filter the dataset by the date range
filtered_states_data_2 <- filtered_states_data %>%
  filter(date >= start_date & date <= end_date) %>% # 17160 by 7
  filter(!state %in% c("Puerto Rico", "Virgin Islands", "Guam", "Northern Mariana Islands", "American Samoa")) %>%
  arrange(date)

dim(filtered_states_data_2) # [1] 16830    7

# Filter to retain the last recorded date per month for each state
monthly_state_data2 <- filtered_states_data_2 %>%
  rename(State = state) %>%
  mutate(year = year(date),                      # Extract the year from the date  
         month = month(date, label = TRUE)) %>%  # Extract the month as a label
  group_by(State, year, month) %>%               # Group by state, year, and month
  filter(date == max(date)) %>%                  # Keep rows where the date is the maximum within each group
  ungroup() %>%
  select(date, month, year, everything())       # Include all columns but exclude the `date` column


# Convert full state names in monthly_us_state_data to state abbreviations
monthly_state_data2$State<- state2abbr(monthly_state_data2$State)

# Convert full state names in population_data to state abbreviations
population_data$State<- state2abbr(population_data$State)

```


Secondly, the `monthly_state_data2` dataset was joined with the `monthly_racial_data` and `population_data`, to create the `combined_monthly_data` that has the integration of COVID-19 racial statistics with population metrics. 


```{r}
# Adjust dates in monthly_racial_data to the end of the month
monthly_racial_data <- monthly_racial_data %>%
  mutate(Date = ceiling_date(Date, "month") - days(1))

# Rename the column 'Date' in monthly_racial_data to match monthly_state_data2
colnames(monthly_racial_data)[colnames(monthly_racial_data) == "Date"] <- "date"

# Adjust dates in monthly_state_data2 to the end of the month
monthly_state_data2 <- monthly_state_data2 %>%
  mutate(date = ceiling_date(date, "month") - days(1))

# Join the datasets
combined_monthly_data <- monthly_state_data2 %>%
  left_join(monthly_racial_data, by = c("date", "month", "year", "State")) %>%
  left_join(population_data, by = c("State", "Population_Total"))

```

Thirdly, in the `combined_monthly_data`, Rates for cases, deaths, hospitalizations, and tests were calculated for each racial/ethnic group by dividing the respective counts by the population of that group and scaling to 100,000. These standardized rates facilitate meaningful comparisons across states and demographic groups.

```{r}
# Calculate rates for columns 13 to 44 based on population values from columns 45 to 52, rounded to 8 decimal places
combined_monthly_data <- combined_monthly_data %>%
  mutate(
    across(
      .cols = 10:41,  # Columns for cases, deaths, hospitalizations, tests
      .fns = ~ round((. / get(sub("Cases|Deaths|Hospitalized|Tests", "Population", cur_column())) * 100000), 8),
      .names = "{.col}_rate"
    )
  )

```

The cleaned dataset, `combined_monthly_data`, includes variables for monthly cases, deaths, hospitalizations, and tests for various racial and ethnic groups, along with population data and calculated rates. Below is a preview of the dataset structure:


Below is a preview of the cleaned `combined_monthly_data` dataset:

```{r}
datatable(combined_monthly_data, filter = "top")
```

<div style="text-align: center; margin-top: 20px;">
  <a href="#top" style="text-decoration: none; background-color: #007BFF; color: white; padding: 10px 20px; border-radius: 5px; font-size: 16px;">Back to Top</a>
</div>


# Exploratory Data Analysis

To Do List [Delete once done]

  - 4.1 Uncover new information in the data that is not self-evident (i.e. do not just plot the data as it is; rather, slice and dice the data in different ways, create new variables, or join separate data frames to create new summary information). 
  - 4.2 Provide findings in the form of plots and tables. Show me you can display findings in different ways. 
  - 4.3 Graph(s) are carefully tuned for desired purpose. One graph illustrates one primary point and is appropriately formatted (plot and axis titles, legend if necessary, scales are appropriate, appropriate geoms used, etc.). 
  - 4.4 Table(s) carefully constructed to make it easy to perform important comparisons. Careful styling highlights important features. Size of table is appropriate. 
  - 4.5 Insights obtained from the analysis are thoroughly, yet succinctly, explained. Easy to see and understand the interesting findings that you uncovered. 

## National-Level Data Exploration


```{r, fig.height=6, fig.width=8}
# Define major U.S. COVID-19 events
us_events <- data.frame(
  date = as.Date(c("2020-03-13", "2020-12-14", "2021-06-15", 
                   "2022-01-03", "2022-09-01", "2023-01-30")),
  event = c("U.S. Declares National Emergency", 
            "First Vaccine Administered", 
            "Delta Variant   Surge Begins",
            "Omicron Surge Peaks", 
            "Bivalent Boosters Approved",
            "End of COVID-19 Emergencies"))


# Plot weekly deaths with vertical annotations

ggplot(monthly_us_data, aes(x = date, y = cases)) +
  geom_line(color = "blue", size = 1) +
  # Add vertical dashed lines for major events
  geom_vline(data = us_events, aes(xintercept = date), color = "magenta", linetype = "dashed", size = 1) +
  # Add vertically rotated text annotations for events
  geom_text(data = us_events, aes(x = date, y = 25000000, label = event), 
            angle = 90, hjust = 0, vjust = -0.5, size = 3, color = "magenta", face = "bold") +
  # Y-axis formatting
  scale_y_continuous(labels = scales::comma) +
  
  # X-axis formatting 
  scale_x_date(date_labels = "%b %Y", date_breaks = "4 months") +
  
  # Add plot titles and labels
  labs(title = "Monthly Cumulative COVID-19 Cases in the U.S.",
    subtitle = "Major COVID-19 Events Highlighted (2020-2023)",
    x = "Date",
    y = "Monthly Cases",
    caption = "Source: National-level Data | Events: Major U.S. Milestones") +
  
  # Minimal theme 
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    panel.grid.major.y = element_line(color = "gray90"))

```


`
```{r, fig.height=6, fig.width=10}
# Plot weekly deaths with vertical annotations

ggplot(monthly_us_data, aes(x = date, y = deaths)) +
  geom_line(color = "red", size = 1) +
  # Add vertical dashed lines for major events
  geom_vline(data = us_events, aes(xintercept = date), color = "blue", linetype = "dashed", size = 1) +
  # Add vertically rotated text annotations for events
  geom_text(data = us_events, aes(x = date, y = 400000, label = event), 
            angle = 90, hjust = 0, vjust = -0.5, size = 3, color = "blue") +
  # Y-axis formatting
  scale_y_continuous(labels = scales::comma) +
  
  # X-axis formatting 
  scale_x_date(date_labels = "%b %Y", date_breaks = "4 months") +
  
  # Add plot titles and labels
  labs(title = "Monthly Cumulative COVID-19 Deaths in the U.S.",
    subtitle = "Major COVID-19 Events Highlighted (2020-2023)",
    x = "Date",
    y = "Monthly Deaths",
    caption = "Source: National-level Data | Events: Major U.S. Milestones") +
  
  # Minimal theme 
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 10),
    panel.grid.major.y = element_line(color = "gray90"))

```



```{r, fig.height=8, fig.width=10}
# Scale the death rate
monthly_us_data <- monthly_us_data %>%
  mutate(deaths_rate_scaled = us_deaths_rate * 50)  # Scale death rate by a factor of 50 for visibility

# Create the plot
ggplot(monthly_us_data, aes(x = date)) +
  # Area plot for Case Rate
  geom_area(aes(y = us_cases_rate, fill = "Case Rate")) +
  # Area plot for Scaled Death Rate
  geom_area(aes(y = deaths_rate_scaled, fill = "Death Rate (Scaled x50)")) +
  # Custom Fill Colors
  scale_fill_manual(values = c("Case Rate" = "blue", "Death Rate (Scaled x50)" = "red"),
    name = "") +
  # Labels and titles
  labs(title = "Monthly Cumulative Rates of COVID-19 Cases and Deaths (per 100,000 Population)",
       subtitle = "Death Rate Scaled for Visual Comparison (x50)",
       x = "Date",
       y = "Rate per 100,000",
       fill = "Metric") +
  # Format the x-axis
  scale_x_date(date_labels = "%b %Y", date_breaks = "3 months") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom")


```


## State-Level Data Exploration


```{r}
# Step 1: Filter and Prepare the Data
subset1 <- monthly_state_data %>%
  filter(month == "Mar" & year == 2023)        # Filter for March 2023 since data is cumulative

# Convert state names to abbreviations
subset1$state <- state2abbr(subset1$state)     # Ensure 'state' has two-letter state abbreviations

# Replace any NA values in the 'state' column with "PR" (Puerto Rico) to ensure all states and territories are properly labeled.
subset1$state[is.na(subset1$state)] <- "PR"


# Step 2: Create Hover Text
subset1$hover_text <- paste0("State: ", subset1$state, "<br>",    # State abbreviation
  "Total Cases: ", scales::comma(subset1$cases), "<br>",          # Total number of cases with commas
  "Cases per 100k: ", round(subset1$state_cases_rate, 4), "<br>", # Cases per 100,000 rounded to 4 decimals
  "Total Deaths: ", scales::comma(subset1$deaths), "<br>",        # Total number of deaths with commas
  "Deaths per 100k: ", round(subset1$state_deaths_rate, 4))       # Deaths per 100,000 rounded to 4 decimals


# Step 3: Create an Interactive US Map Using plot_geo
fig1 <- plot_geo(subset1, locationmode = 'USA-states') %>%        # Create a US map with state boundaries
  add_trace(z = ~ cases,          # Value to display (total cases)
            locations = ~ state,  # State abbreviations for mapping
            text = ~ hover_text,  # Custom hover text
            hoverinfo = "text",   # Display only custom hover text
            colorscale = "Reds",  # Color gradient for total cases (light to dark red)
            colorbar = list(title = "Total Cases")) %>% # Add color bar legend
  layout(title = list(
      text = "U.S. State Map of Cumulative COVID-19 Cases (March 23, 2023)<br><sup>Hover over the map to see more details</sup>",
      x = 0.5, # Center the title
      font = list(size = 18)),
      geo = list(scope = 'usa',                          # Focus map on the USA
                 projection = list(type = 'albers usa'), # Use Albers USA map projection
                 showlakes = TRUE,                       # Show lakes on the map
                 lakecolor = 'rgb(255, 255, 255)'))      # Set lake color to white

# Step 4: Display the Interactive Map
fig1 
```


```{r, fig.height=10, fig.width=15}
# Visualizing number of Covid cases per state
plot1 <- ggplot(data = subset1) +
  geom_bar(aes(x = reorder(state, cases),
               y = cases, fill = state),
           stat = "identity", show.legend = FALSE) +
  scale_y_continuous(labels = label_comma()) + # Full number format
  labs(x = "State", y = "Covid Cases", title = "Cumulative Covid Cases per State (March 23, 2023)") +
  coord_flip() +
  theme_classic() +
  theme(plot.margin = margin(t = 10, r = 20, b = 20, l = 50), # Adjust margins
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.text.y = element_text(size = 14, face = "bold")) # Increase font size and bold state names
  

# Visualizing number of Covid deaths per state
plot2 <- ggplot(data = subset1) +
  geom_bar(aes(x = reorder(state, deaths),
               y = deaths, fill = state),
           stat = "identity", show.legend = FALSE) +
  labs(x = "State", y = "Covid Deaths", title = "Cumulative Covid Deaths per State (March 23, 2023)") +
  coord_flip() +
  theme_classic() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),,
        axis.text.y = element_text(size = 12, face = "bold")) # Increase font size and bold state names

# Arrange both plots side by side
grid.arrange(plot1, plot2, ncol = 2)

```




## Covid Racial Data Exploration


### Reported Cases and Deaths

```{r}
# Aggregate data by month and year
monthly_racial_cases <- combined_monthly_data %>%
  group_by(date) %>%
  summarize(Cases_White = sum(Cases_White, na.rm = TRUE),
            Cases_Black = sum(Cases_Black, na.rm = TRUE),
            Cases_Asian = sum(Cases_Asian, na.rm = TRUE),
            Cases_AIAN = sum(Cases_AIAN, na.rm = TRUE),
            Cases_NHPI = sum(Cases_NHPI, na.rm = TRUE),
            Cases_Multiracial = sum(Cases_Multiracial, na.rm = TRUE),
            .groups = "drop") # Removes grouping

# Display the aggregated monthly data
monthly_racial_cases %>%
  kbl(caption = "Cumulative Monthly COVID-19 Cases by Racial Group", 
      col.names = c("Date", "White", "Black", "Asian", "AIAN", "NHPI", "Multiracial"),
      format = "html") %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                position = "center") %>%
  column_spec(1, bold = TRUE, width = "8em") %>%   # Bold and narrow Date column
  column_spec(2:7, width = "8em") %>%               # Set consistent column width
  row_spec(0, bold = TRUE, background = "magenta")  # Highlight the header row

```


```{r}
# Aggregate deaths data by month
monthly_racial_deaths <- combined_monthly_data %>%
  group_by(date) %>%
  summarize(
    Deaths_White = sum(Deaths_White, na.rm = TRUE),
    Deaths_Black = sum(Deaths_Black, na.rm = TRUE),
    Deaths_Asian = sum(Deaths_Asian, na.rm = TRUE),
    Deaths_AIAN = sum(Deaths_AIAN, na.rm = TRUE),
    Deaths_NHPI = sum(Deaths_NHPI, na.rm = TRUE),
    Deaths_Multiracial = sum(Deaths_Multiracial, na.rm = TRUE),
    .groups = "drop"
  )

# Display the aggregated monthly data
monthly_racial_deaths %>%
  kbl(caption = "Cumulative Monthly COVID-19 Deaths by Racial Group", 
      col.names = c("Date", "White", "Black", "Asian", "AIAN", "NHPI", "Multiracial"),
      format = "html") %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                position = "center") %>%
  column_spec(1, bold = TRUE, width = "8em") %>%    # Bold and narrow Date column
  column_spec(2:7, width = "8em") %>%               # Set consistent column width
  row_spec(0, bold = TRUE, background = "magenta")  # Highlight the header row
```

It is essential to plot **rates** instead of raw counts when analyzing public health metrics like COVID-19 cases, deaths, hospitalizations, and testing because rates provide a more accurate and fair comparison across groups. Different racial or demographic groups have varying population sizes, and raw counts can misrepresent the true burden of disease if one group is significantly larger than another. By normalizing counts to a standard unit (e.g., per 100,000 people), rates adjust for population size differences, ensuring meaningful and equitable comparisons. Thus, for the next section, we compare rates of cases, deaths, hospitalizations and tests by racial groups.

### Rates of Cases and Deaths


```{r}
# Aggregate data by month for case rates
monthly_racial_case_rate <- combined_monthly_data %>%
  group_by(date) %>%
  summarize(
    Cases_White = sum(Cases_White_rate, na.rm = TRUE),
    Cases_Black = sum(Cases_Black_rate, na.rm = TRUE),
    Cases_Asian = sum(Cases_Asian_rate, na.rm = TRUE),
    Cases_AIAN = sum(Cases_AIAN_rate, na.rm = TRUE),
    Cases_NHPI = sum(Cases_NHPI_rate, na.rm = TRUE),
    Cases_Multiracial = sum(Cases_Multiracial_rate, na.rm = TRUE),
    .groups = "drop")  # Remove grouping
  

# Display the aggregated death rate table
monthly_racial_case_rate %>%
  kbl(caption = "Cumulative Monthly Rates of COVID-19 Cases by Racial Group", 
      col.names = c("Date", "White", "Black", "Asian", "AIAN", "NHPI", "Multiracial"),
      format = "html",  
      digits = 4) %>%   
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                position = "center") %>%
  column_spec(1, bold = TRUE, width = "8em") %>%   # Bold and narrow Date column
  column_spec(2:7, width = "8em") %>%               # Set consistent column width
  row_spec(0, bold = TRUE, background = "magenta")  # Highlight the header row

```

```{r, fig.width=8, fig.height=6}
# Reshape data to long format for plotting
monthly_racial_case_rate_long <- monthly_racial_case_rate %>%
  pivot_longer(cols = c(Cases_White, Cases_Black, Cases_Asian, Cases_AIAN, 
                        Cases_NHPI, Cases_Multiracial),
               names_to = "racial_group",
               values_to = "case_rate") %>%
  mutate(racial_group = str_remove(racial_group, "Cases_")) %>%  # Remove "Cases_" prefix
  arrange(date, case_rate)

# Define breaks and date range 
date_breaks <- seq(min(monthly_racial_case_rate$date), 
                   max(monthly_racial_case_rate$date), by = "1 month")

# Custom colors for racial groups
custom_colors <- c("White" = "magenta", "Black" = "black",
  "Asian" = "orange", "AIAN" = "brown", "NHPI" = "blue", "Multiracial" = "green")


# Plot the case rates
ggplot(monthly_racial_case_rate_long, aes(x = date, y = case_rate, group = racial_group, color = racial_group)) +
  geom_line(size = 1) +         
  geom_point(size = 2) +       
  scale_color_manual(values = custom_colors) +
  scale_x_date(date_labels = "%m/%Y", breaks = date_breaks, expand = c(0.01, 0.01)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Cumulative Monthly Rates of COVID-19 Cases by Racial Group",
       x = "Month/Year",
       y = "Case Rate (per 100,000)",
       color = "Racial Group") +
  theme_minimal() +
    theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "top")

```


```{r}
# Aggregate data by month for death rates
monthly_racial_death_rate <- combined_monthly_data %>%
  group_by(date) %>%
  summarize(
    Deaths_White = sum(Deaths_White_rate, na.rm = TRUE),
    Deaths_Black = sum(Deaths_Black_rate, na.rm = TRUE),
    Deaths_Asian = sum(Deaths_Asian_rate, na.rm = TRUE),
    Deaths_AIAN = sum(Deaths_AIAN_rate, na.rm = TRUE),
    Deaths_NHPI = sum(Deaths_NHPI_rate, na.rm = TRUE),
    Deaths_Multiracial = sum(Deaths_Multiracial_rate, na.rm = TRUE),
    .groups = "drop")  # Removes grouping
  

# Display the aggregated death rate table
monthly_racial_death_rate%>%
  kbl(caption = "Cumulative Monthly Rates of COVID-19 Deaths by Racial Group", 
      col.names = c("Date", "White", "Black", "Asian", "AIAN", "NHPI", "Multiracial"),
      format = "html",  
      digits = 4) %>%   
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                position = "center") %>%
  column_spec(1, bold = TRUE, width = "8em") %>%   # Bold and narrow Date column
  column_spec(2:7, width = "8em") %>%               # Set consistent column width
  row_spec(0, bold = TRUE, background = "magenta")  # Highlight the header row
```


```{r, fig.width=8, fig.height=6}
# Reshape data to long format for plotting
monthly_racial_death_rate_long <- monthly_racial_death_rate %>%
  pivot_longer(cols = c(Deaths_White, Deaths_Black, Deaths_Asian, Deaths_AIAN, 
                        Deaths_NHPI, Deaths_Multiracial),
               names_to = "racial_group",
               values_to = "death_rate") %>%
  mutate(racial_group = str_remove(racial_group, "Deaths_")) %>% # Clean column names
  arrange(date, death_rate)

# Plot the data
ggplot(monthly_racial_death_rate_long, aes(x = date, y = death_rate, group = racial_group, color = racial_group)) +
  geom_line(size = 1) +         
  geom_point(size = 2) +       
  scale_color_manual(values = custom_colors) +
  scale_x_date(date_labels = "%m/%Y", breaks = date_breaks, expand = c(0.01, 0.01)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Cumulative Monthly Rates of COVID-19 Deaths by Racial Group",
       x = "Month/Year",
       y = "Death Rate (per 100,000)",
       color = "Racial Group") +
  theme_minimal() +
    theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "top")

```


### Rates of Hospitalizations and Tests

```{r, fig.width=9, fig.height=7}
# Aggregate hospitalized rates data by month
monthly_racial_hospitalized <- combined_monthly_data %>%
  group_by(date) %>%
  summarize(Hospitalized_White = sum(Hospitalized_White_rate, na.rm = TRUE),
            Hospitalized_Black = sum(Hospitalized_Black_rate, na.rm = TRUE),
            Hospitalized_Asian = sum(Hospitalized_Asian_rate, na.rm = TRUE),
            Hospitalized_AIAN = sum(Hospitalized_AIAN_rate, na.rm = TRUE),
            Hospitalized_NHPI = sum(Hospitalized_NHPI_rate, na.rm = TRUE),
            Hospitalized_Multiracial = sum(Hospitalized_Multiracial_rate, na.rm = TRUE),
            .groups = "drop")

# Reshape data into long format for hospitalized rates
monthly_racial_hospitalized_long <- monthly_racial_hospitalized %>%
  pivot_longer(cols = starts_with("Hospitalized_"),
    names_to = "racial_group",
    values_to = "hospitalized_rate") %>%
  mutate(racial_group = str_remove(racial_group, "Hospitalized_"))  # Clean column names

# Plot the data
ggplot(monthly_racial_hospitalized_long, aes(x = date, y = hospitalized_rate, group = racial_group, color = racial_group)) +
  geom_line(size = 1) +         
  geom_point(size = 2) +       
  scale_color_manual(values = custom_colors) +
  scale_x_date(date_labels = "%m/%Y", breaks = date_breaks, expand = c(0.01, 0.01)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Cumulative Monthly Rates of COVID-19 Hospitalizations by Racial Group",
       x = "Month/Year",
       y = "Hospitalization Rates (per 100,000)",
       color = "Racial Group") +
  theme_minimal() +
    theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "top")
```

```{r, fig.width=8, fig.height=6}
# Aggregate test rates data by month
monthly_racial_tests <- combined_monthly_data %>%
  group_by(date) %>%
  summarize(Tests_White = sum(Tests_White_rate, na.rm = TRUE),
            Tests_Black = sum(Tests_Black_rate, na.rm = TRUE),
            Tests_Asian = sum(Tests_Asian_rate, na.rm = TRUE),
            Tests_AIAN = sum(Tests_AIAN_rate, na.rm = TRUE),
            Tests_NHPI = sum(Tests_NHPI_rate, na.rm = TRUE),
            Tests_Multiracial = sum(Tests_Multiracial_rate, na.rm = TRUE),
            .groups = "drop")

# Reshape data into long format for test rates
monthly_racial_tests_long <- monthly_racial_tests %>%
  pivot_longer(cols = starts_with("Tests_"),
               names_to = "racial_group",
               values_to = "test_rate") %>%
  mutate(racial_group = str_remove(racial_group, "Tests_"))  # Clean column names

# Plot the data
ggplot(monthly_racial_tests_long, aes(x = date, y = test_rate, group = racial_group, color = racial_group)) +
  geom_line(size = 1) +         
  geom_point(size = 2) +       
  scale_color_manual(values = custom_colors) +
  scale_x_date(date_labels = "%m/%Y", breaks = date_breaks, expand = c(0.01, 0.01)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Cumulative Monthly Rates of COVID-19 Tests by Racial Group",
       x = "Month/Year",
       y = "Tests Rate (per 100,000)",
       color = "Racial Group") +
  theme_minimal() +
    theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    legend.position = "top")
```


<blockquote style="font-size: 1.3em; font-style: italic; color: #555555; margin-top: 20px; margin-bottom: 20px; border-left: 4px solid #cccccc; padding-left: 15px;">
“What are the relationships between COVID-19 case rates and key predictors: testing rates, hospitalization rates, and death rates across different racial groups, and do these predictors explain disparities in case rates among racial groups over time?”
</blockquote>


Finally, in this section, we conduct a linear regression analysis to identify and quantify the relationships between the **COVID-19 case rates** (dependent variable) and key predictors (independent variables) across different **racial groups** over time. The goal of this model is to understand:

1. **How Testing Rates, Hospitalization Rates, and Death Rates Relate to Case Rates**:
   - Does an increase in testing rates correlate with higher case rates?  
   - Are hospitalization rates or death rates indicators of the severity or extent of infections?  
2. **Differences Across Racial Groups**:
   - By including `racial_group` as a categorical variable, the model accounts for differences in case rates across racial groups.  
   - This allows us to see if certain racial groups are more impacted when adjusting for predictors like testing, hospitalization, and death rates.  

**Model Specification**

The linear regression model is specified as follows:

$$
\text{Case Rate} = \beta_0 + \beta_1 (\text{Testing Rate}) + \beta_2 (\text{Hospitalized Rate}) + \beta_3 (\text{Death Rate}) + \beta_4 (\text{Racial Group}) + \epsilon
$$

```{r}
# Reshape the data and keep only relevant variables
combined_long <- combined_monthly_data %>%
  # Reshape the necessary columns into long format
  pivot_longer(cols = c(Cases_White_rate, Cases_Black_rate, Cases_Asian_rate, Cases_AIAN_rate, 
             Cases_NHPI_rate, Cases_Multiracial_rate,
             Tests_White_rate, Tests_Black_rate, Tests_Asian_rate, Tests_AIAN_rate, 
             Tests_NHPI_rate, Tests_Multiracial_rate,
             Hospitalized_White_rate, Hospitalized_Black_rate, Hospitalized_Asian_rate, 
             Hospitalized_AIAN_rate, Hospitalized_NHPI_rate, Hospitalized_Multiracial_rate,
             Deaths_White_rate, Deaths_Black_rate, Deaths_Asian_rate, Deaths_AIAN_rate, 
             Deaths_NHPI_rate, Deaths_Multiracial_rate),
    names_to = c("variable_type", "racial_group"),  # Separate variable name and racial group
    names_pattern = "(.*)_(.*)_rate",              # Split column names into parts
    values_to = "value") %>%
  # Reshape back into wide format keeping only the relevant predictors
  pivot_wider(
    names_from = variable_type,     # Spread variable types into separate columns
    values_from = value) %>%             # Values go into the respective columns
  # Select only the needed variables for linear regression
  select(date, racial_group, Cases, Tests, Hospitalized, Deaths)


# Run linear regression
model_combined <- lm(Cases ~ Tests + Hospitalized + Deaths + racial_group, data = combined_long)

# Tidy the regression model coefficients
model_results <- tidy(model_combined)

# Rename predictors for better readability
model_results <- model_results %>%
  mutate(term = case_when(
      term == "(Intercept)" ~ "Intercept",
      term == "racial_groupAsian" ~ "Asian Group",
      term == "racial_groupBlack" ~ "Black Group",
      term == "racial_groupMultiracial" ~ "Multiracial Group",
      term == "racial_groupNHPI" ~ "NHPI Group",
      term == "racial_groupWhite" ~ "White Group",
      term == "Tests" ~ "Testing Rate",
      term == "Hospitalized" ~ "Hospitalization Rate",
      term == "Deaths" ~ "Death Rate",
      TRUE ~ term),
    Estimate = round(estimate, 3),
    `Std. Error` = round(std.error, 3),
    `t value` = round(statistic, 3),
    `p value` = ifelse(p.value < 0.001, "<0.001", round(p.value, 3)),
    Stars = case_when(   # Add stars based on p-value thresholds
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      p.value < 0.1   ~ ".",
      TRUE            ~ "")) %>%
  mutate(`p value` = paste0(`p value`, Stars)) %>%  # Append stars to p-values
  select(term, Estimate, `Std. Error`, `t value`, `p value`)

# Extract model statistics using glance()
model_stats <- glance(model_combined)

# Combine model statistics into a footnote string
model_footnote <- paste0(
  "Residual Standard Error: ", round(model_stats$sigma, 3), 
  " | Multiple R-squared: ", round(model_stats$r.squared, 4), 
  " | Adjusted R-squared: ", round(model_stats$adj.r.squared, 4),
  " | F-statistic: ", round(model_stats$statistic, 2), 
  " on ", model_stats$df, " and ", model_stats$df.residual, " DF",
  " | p-value: < 2.2e-16"
)

# Display the coefficient table with footnotes
model_results %>%
  kbl(caption = "Regression Results: COVID-19 Case Rates",
      col.names = c("Predictor", "Estimate", "Std. Error", "t Value", "p Value"),
      format = "html") %>%
  kable_styling(full_width = FALSE, 
                bootstrap_options = c("striped", "hover", "condensed", "responsive"), 
                position = "center") %>%
  row_spec(0, bold = TRUE, background = "magenta") %>%
  column_spec(1, bold = TRUE) %>%
  footnote(general = paste(model_footnote, 
                           "Signif. codes: 0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1"))

```


The results of the linear regression model demonstrate that **testing rates**, **hospitalization rates**, and **death rates** are significant predictors of COVID-19 case rates, explaining over **94% of the variability** (Adjusted R-squared = 0.9416). A positive relationship was observed between testing rates and case rates, suggesting that increased testing identifies more COVID-19 cases, as expected. Testing is critical for case detection, particularly in underrepresented communities.. Similarly, higher hospitalization and death rates were associated with increased case rates. This strong and positive association may reflect the severity of COVID-19 cases where higher hospitalization rates correlate with increased case identification. It could also reflect a lagging relationship, where severe illness leads to more comprehensive testing.
The significant association between death rates and cases rates underscores that higher death rates often occur alongside higher case rates. This trend could reflect regions with higher COVID-19 transmission or communities with limited access to early interventions.


The racial group predictors estimate how COVID-19 case rates differ among racial groups relative to a reference category (AIAN).  Compared to the AIAN, **Multiracial**, **Asian**, and **Black communities** experienced notably higher case rates, possibly indicating systemic inequities that may stem from limited access to healthcare, socioeconomic vulnerabilities, and other social determinants of health.  While the **White** racial group had significant higher case rates,the disparity here may not be as pronounced as in some other racial groups. The **NHPI group** showed a lower case rate. This result contrasts with trends seen in other groups and may reflect differences in reporting, access to testing, or population size considerations.





# Summary

To Do List [Delete Once Done]

  - 6.1 Summarize the problem statement you addressed. 
  - 6.2 Summarize how you addressed this problem statement (the data used and the methodology employed). 
  - 6.3 Summarize the interesting insights that your analysis provided. 
  - 6.4 Summarize the implications to the reader of your analysis. 
  - 6.5 Discuss the limitations of your analysis and how you, or someone else, could improve or build on it. 



# References

<div id= "refs"></div>


:::
